---
title: "Homework"
subtitle: 'Data frames'
author:
  - name: Mikołaj Biesaga 
    url: http://iss.uw.edu.pl/mikolaj-biesaga
    affiliation: Robert Zajonc Institute for Social Studies, University of Warsaw
    affiliation_url: http://iss.uw.edu.pl
output:
   distill::distill_article:
     highlight: haddock
     code_folding: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = TRUE,
  message = TRUE,
  comment = "#",
  R.options = list(width = 60)
)
```

## Introduction

We are aware of the fact that these tasks might be difficult. We do not expect
you to know the answer right away. First, you should try to review the
`SocEli-0.Rmd` and the beginning of the `SocEli-1.Rmd` we covered on Friday. And
afterward, just try to look for the solution online. It might be either in the
books we recommend to you or just on StackOverflow.

```{r loadLibraries, include=FALSE}
library(tidyverse) ## every time we open RStudio we should load 
                   ## libraries we are going to use
```

## Task 1

From the `iris` data set extract column `species` and assign it to the object
called `species_v`. Create a new vector that instead of setosa would have red,
versicolor -- green, and virginica -- blue.


```{r recodeIris}
## WRITE YOUR ANSWER HERE

## Extract Species
iris_v <- iris$Species
## Create a lookup vector
lookup_v <- c('setosa' = 'red', 'versicolor' = 'green', 'virginica' = 'blue')
## New vector
new_v <- lookup_v[iris_v]
```

## Task 2

Imagine that you have a data set with two columns. The first indicates whether
the person got the first dose of coronavirus vaccine and the second if got a
second jab. You would like to talk to all people who were vaccinated at least
with one doze. You don't care whether it was the first or second jab, or did
they are fully vaccinated or not. Count the number of interviews you are meant
to schedule.

```{r vaccination}
set.seed(42) ## this is important because below we use a pseudorandomization 
## function -- sample(). It draws in random order 100 times from a vector 
## c(TRUE, FALSE, NA) with replacement. We want to make sure that for all 
## of us it will draw in exactly the same order. Therefore, we tell R that 
## it should use a build-in Random Number Generation in the specific state 
## (the number we put as an argument in the function does not matter while 
## we all have the same). Try commenting out this function and running the 
## code a few times. What is the difference?

## Data set
df_vaccination <- tibble( 'first' = sample(c(TRUE,FALSE), 100, replace = TRUE),
                          'second' = sample(c(TRUE, FALSE, NA), 100, replace = TRUE))

## WRITE YOUR ANSWER HERE
df_vaccination %>%
  filter(first | second) %>%
  nrow()
```

## Task 3

Now you have similar data set as in the previous task but you would like to
understand why people did not take the vaccines. Count the number of interviews.

```{r vaccinationRefuse}
set.seed(92210) ## Let's not make it trivial and change the data a bit

## Data set
df_vaccination <- tibble( 'first' = sample(c(TRUE,FALSE), 100, replace = TRUE),
                          'second' = sample(c(TRUE, FALSE, NA), 100, replace = TRUE))

## WRITE YOUR ANSWER HERE
df_vaccination %>%
  filter(!(first) & (!(second) | is.na(second) )) %>%
  nrow()
```

## Task 4

Imagine that you are the Minister of Education in Poland (trust me for sure you
are qualified and being fluent in Polish is not necessary). One day you got the
result of the final exam in Math every student in Poland takes at the end of
high school. You would like to know how many people got more than 90 points out
of 100.

```{r vectorCount}
set.seed(314)
## A very long vector with results
vector_c <- sample(c(1:100), 100000, replace = TRUE)

## WRITE YOUR ANSWER HERE
sum(vector_c > 90)
```