---
title: "Introduction to igraph"
author: "Szymon Talaga and Mikołaj Biesaga"
affiliation: "Robert Zajonc Institute for Social Studies"
output:
  reprex::reprex_document:
    venue: "gh"
    advertise: FALSE
    style: TRUE
    tidyverse_quiet: TRUE
    std_out_err: TRUE
    session_info: TRUE
    comment: "#"
knit: reprex::reprex_render
---
## Introduction

We have already covered some basics of `R` and more or less you understand the
workflow in `RStudio`. I mean, I tried to show you these two things but it took
me a lot of practice (and frustration) to be able to say that I more or less
know what I am doing. So don't worry if you still feel uneasy with all that.
Believe me. It is really just a matter of practice.

In this notebook, we are finally going to talk about networks. These are going
to be very basics because Szymon will have the whole part about it. The aim of
this notebook is just to show you some simple visualizations of networks in R,
and simple computations you can perform. To do that we are going to use `igraph`
package.

## Load Packages

Let's do it a bit differently this time. The concept is really easy. We want to
first check whether the package we want to use is already installed on our local
machine. If it is installed we want to load it but if not we want to first
install it. How to do it in `R`? Obviously, there are multiple ways how to
implement it but let's focus on the scheme, not on the code. The scheme is as
follows.

1. We need a list of packages we want to use in the notebook. It might be a list
but a character vector would be good enough.
2. We take the first element from the list (vector) -- the name of the package.
3. We check whether this package is already installed.
4. If it is not installed we simply install it.
5. Now, regardless of whether it had been installed previously or we just
installed it, we have the package in question on our local machine. Therefore,
we can install it.

After step number 5 we take the next package from our list (vector) until we
reach the end of it. This is quite a simple algorithm that you would figure out
yourself. In programming languages, it is a simple example of a for-loop
combined with an if-statement.

```{r loadLibraries}
## Create a vector with names of the packages we want to load
packages <- c('tidyverse', 'igraph', 'ggraph')

## Instead of a for-loop we are going to use a function lapply.
## It is extremely powerful and useful. It takes a list (or a vector
## in that case) and applies a function to each of its elements.
## Therefore, it takes two arguments: 1. list (or a vector)
## 2. a function we want to apply. It returns always a list.
## In our case we are going to define a simple function ourselves
## It will first check whether the package in question is in 
## among installed packages. If it is not, it will install it.
## Afterwards it will load the package. We will use a function
## require() instead of library() because inside a function
## it is impossible to use library(). You can try to change it
## and see what happens.
lapply(packages, function(package){
  if (!(package %in% installed.packages())){
    install.packages(package)
  }
  library(package, character.only = TRUE, logical.return = TRUE)
})
```

## Networks in `igraph`

Before we start with creating our first networks let's have a common
understanding of some basic concepts.

* A **network** refers to a structure representing a group of
objects/actors/people and the relationship between them. In mathematics, it is
known under the name graph. A network consists of nodes and edges.
* **Node**, also vertex, represents an actor/object. It might have various
properties that can be displayed in a network using its size, color, shape, etc.
* **Edge**, also link, depicts the relationship between actors/objects. In a
simple case edges represent the existence of the relationship, therefore, either
there is a link between two objects or not. However, in more advanced cases the
link might have the property of direction (directed networks) or intensity
(weighted networks).

Before we talk about how to create networks in `R`, let's think about what kind
of data we need. In general, the data to create a network might come in two
forms:

1. Edges list and nodes list. In some cases a list of edges is sufficient but to
be 100% sure that we capture all nodes we better have both lists, for example:
2. Adjacency Matrix. It is a square matrix. The elements of the matrix indicate
whether pairs of nodes (vertexes) are adjacent (are in the relationship) or not,
for example:

### Undirected Networks

The simplest network we can draw is an undirected one. It means that the nodes
are connected with links that do not indicate the direction of the relationship.
In terms of social science, you might think about the below network as a
representation of friends on Facebook among 5 people. Being friends on Facebook
does not have direction. You either are friends or not.

The simplest way to create a network in `igraph` is by defining a vector of
edges. The issue is, however, that it not that easy as it was in the example
above. The vector of edges in `igraph` takes the following scheme: $
c(n_{1from}, n_{1to}, n_{2from}, n_{2to},...)$. Therefore, you need to be very
careful when you construct such a vector. You also probably would like to
specify the number of nodes. That way you indicate whether all nodes have at
least one edge or there are some isolated nodes.

```{r undirected}
## Create a igrpah object
g1 <- graph( edges=c(1,2, 2,3, 3,1, 2,5, 4,5, 4,1), n=5, directed = FALSE ) 
plot(g1)
```

### Directed network

In directed networks on the other hand the relationship has also direction. In
terms of being friends on Facebook, this graph might depict who sent the friend
request to whom.

```{r directed}
g2 <- graph( edges=c(1,2, 2,3, 3, 1, 2, 5, 4, 5, 4,1), n=5, directed = TRUE )
plot(g2)
```

### Nammed vertexes

In our case most of the time we would like to have named nodes instead of just
numbers. There is an easy way to do it with the use of `igraph`. We can just
create a character vector with the names of nodes. In this case, if we want to
add nodes that are not connected to any other node we put their names as the
value of the isolates argument.

```{r nammedVertexes}
g4 <- graph( c("John", "Jim", "Jim", "Jack", "Jim", "Jack", "John", "John"), 
             isolates=c("Jesse", "Janis", "Jennifer", "Justin"), directed = TRUE )  

plot(g4)
```
### Attributes

Let's now have a closer look at edges and nodes. We have already said that they
might have different attributes that we can visualize using different
aesthetics, i.e. colors, size, shape, etc. However, before we move any further
let's extract information about edges and nodes from our `igraph` object.

```{r edges}
## Extract edges from the object
E(g4)
```

```{r vertexes}
## Extract vertexes from the object
V(g4)
```

We might also want to extract the adjacency matrix of the network. Sometimes it
allows us to better understand what is happening.

```{r adjancyMatrix}
## Extract adjancy matrix
g4[]

## You can extract also sepcific columns or rows by using indexes
g4[1,] ## extract first row
g4[,1] ## extract first column

## This is the adjacency matrix for the directed graph. Would the adjacency matrix for the undirected graph be exactly the same?
```

#### Vertex Attributes

We have already said how vertex attributes might be useful in terms of
visualization. Let's see how we add attributes to vertexes. It is fairly simple.
You just need to add a vector of a number of nodes length or of length 1. In the
latter case, all vertexes would have the same value of a given attribute.

```{r vertexAttributes}
V(g4)$name
V(g4)$weight <- c(1,2,5,4,2,3,7)
V(g4)$gender <- c('male', 'male', 'male', 'female', 'female', 'female', 'male')
vertex_attr(g4)
```

So how to use them in practice. We will use our second package for today for
this -- `ggraph`. It is possible to do it in plain `igraph` but since we learn a
bit of `ggplot` the `ggraph` has a very similar structure. I would say more
straightforward.

```{r vertexAttributesPlot}
## Draw a network with the use of ggraph
g4 %>%
  ## This is very similar to ggplot function
  ggraph(layout = 'kk') +
  ## Simple edge aestethics
  geom_edge_link(aes(start_cap = label_rect(node1.name), end_cap = label_rect(node2.name)), 
                   arrow = arrow(type = "closed", length = unit(3, 'mm'))) +
  ## Loops edge aestethics
  geom_edge_loop(aes(start_cap = label_rect(node1.name), end_cap = label_rect(node2.name)), 
                   arrow = arrow(type = "closed", length = unit(3, 'mm'))) +
  ## Nodes aestethics
  geom_node_point(aes(color = gender, size = weight)) +
  ## Labels aestethics
  geom_node_text(aes(label = name), repel = TRUE) +
  ## Adds defualt theme of the graph (white background)
  theme_graph()
```

One of the most fundamental properties of each node is the so-called degree. It
just means a number of edges that are connected to a given node. There is a
fairly easy way to compute it in `igrpah`.

```{r degree}
## Compute the degree of each node
degree(g4)

## Assign the degree distribution to the object called prob
prob <- degree_distribution(g4, cumulative = FALSE, loop = TRUE, mode = 'total')
## Assign the number of degrees ot the object called degree_n
degree_n <- 0:(length(prob)-1)

## Create a data frame with two collumns degree_n and prob
tibble(degree_n = degree_n,
       prob = prob) %>%
## Plot the degree distribution of the network
  ggplot(aes(x = degree_n, y = prob)) +
  geom_line()  +
  geom_bar(stat = 'identity', alpha = .2 ) +
  theme_classic()
```

```{r degreeExercise}
## Plot degree distribution but without loops and only considering edges that are ingoing the nodes
```

#### Edges Attributes

Similarly to vertexes (nodes') attributes we might add edge attributes to the
network. We can either the same value for all edges or specify them more or less
by hand with a vector of values.

```{r edgeAttributes}
E(g4)$type <- 'email'
E(g4)$weight <- c(.2,.3,.1,.5)
```

Let's see how we can use it in practice. It has a very similar logic to node
attributes because we are going to use the same package -- `ggraph`.

```{r edgeAttributesPlot}
g4 %>%
  ## This is very similar to ggplot function
  ggraph(layout = 'kk') +
  ## Simple edge aestethics
  geom_edge_link(aes(edge_width = weight, start_cap = label_rect(node1.name), end_cap = label_rect(node2.name)), 
                   arrow = arrow(type = "closed", length = unit(3, 'mm')), lintetype = 1) +
  ## Loops edge aestethics
  geom_edge_loop(aes(start_cap = label_rect(node1.name), end_cap = label_rect(node2.name)), 
                   arrow = arrow(type = "closed", length = unit(3, 'mm'))) +
  ## Nodes aestethics
  geom_node_point(aes(color = gender, size = weight)) +
  ## Labels aestethics
  geom_node_text(aes(label = name), repel = TRUE) +
  ## Adds defualt theme of the graph (white background)
  theme_graph() +
  ## Scale the width of edges
  scale_edge_width_continuous(range = c(0.2,1)) +
  ## Remove the legends
  theme(legend.position = 'none')
```

## Exercise

We played a bit with very small artificial networks but in most cases, we will
have bigger networks and we will be not creating the edges in `R`. We will load
the data from the external file (i.e. csv) to `R`. Usually, it will be an edge
list as I mentioned before. How to convert a more traditional `edge_list` to
`igraph` object? We will use for this purpose a build-in data set -- high
school. his dataset shows the friendship among high school boys as assessed by
the question: "What fellows here in school do you go around with most often?".
The question was posed twice, with one year in between (1957 and 1958), and
shows the evolution in friendship between the two time points.

```{r ggraphExerciseExamineData}
## Examine the data
highschool
```


```{r ggraphExerciseCreateGraph}
## Create a graph from edge list
g5 <- graph_from_data_frame(highschool)

## Examine the vertexes
V(g5)

## Examine attributes of vertexes
vertex_attr(g5)

## Examine the edges
E(g5)

## Examine attributes of edges
edge_attr(g5)
```

Now once we have the network. I would like you to try to plot it. I would like
to see a directed network, with colored edges (it should be a function of the
friendship year), and the size of the node should be proportional to the total
degree of the node.

```{r ggraphExrcies}
## WRITE YOUR CODE HERE
```