---
title: "Data manipulation and data visualisation"
subtitle: 'Data frames'
author: "Szymon Talaga and Mikołaj Biesaga"
affiliation: "Robert Zajonc Institute for Social Studies"
output:
  reprex::reprex_document:
    venue: "gh"
    advertise: FALSE
    style: TRUE
    tidyverse_quiet: TRUE
    std_out_err: TRUE
    session_info: TRUE
    comment: "#"
knit: reprex::reprex_render
---

## Introduction

In this notebook, we will focus more on data frames and a tidy approach to data
processing. Therefore, we will use already existing functions and packages.
However, let's first understand what we mean by functions and packages.

* `function` in most programming languages is understood very similarly to a
mathematical function. It is a piece of code that performs a specific task. It
can be called and reused multiple times. We have already used a bunch of
different functions, for example `head()` and `tail()`
* `package` is a code that bundles together functions, data, and documentation
in standardized collection format.

## Load packages

We will use only one external package today. However, regardless of how many you
use the good practice is to load all of them in one of the very first chunks of
code.

```{r loadLibraries}
library(tidyverse) ## load library for data processing
```

## Data manipulation with `dplyr`

Data wrangling or data manipulation is a very time-consuming process. Usually,
when you try to perform some statistical analysis you spend most of the time
trying to either tidy up your data set or organize it in a way that allows you
to perform the analysis you would like to. This is even more true now when on
one hand it is easier to access data than ever before but at the same time
data sets are often messy or not exactly formatted you would like them to be.

We will use here functions from `dplyr` package that aims to provide a function
for each basic verb of data manipulation. These verbs can be organized into
three categories based on the component of the data set that they work with:

* Rows:
  * `filter()` chooses rows based on column values.
  * `arrange()` changes the order of the rows.
  * `slice()` chooses rows based on location.
* Columns:
  * `select()` changes whether or not a column is included.
  * `rename()` changes the name of columns.
  * `mutate()` changes the values of columns and creates new columns.
* Groups of rows:
  * `summarise()` collapses a group into a single row.

### The pipe

All of the `dplyr` functions take a data frame as the first argument. Rather
than forcing the user to either save intermediate objects or nest functions,
`dplyr` provides the `%>%` operator from `magrittr`. `x %>% f(y)` turns into
`f(x, y)` so the result from one step is then “piped” into the next step. You
can use the pipe to rewrite multiple operations that you can read left-to-right,
top-to-bottom (reading the pipe operator as “then”).

### `filter()`

`filter()` allows you to select a subset of rows in a data frame. Like all
single verbs, the first argument is the tibble (or data frame). The second and
subsequent arguments refer to variables within that data frame, selecting rows
where the expression is TRUE.

```{r filter}
## Filter characters from Tatooine with brown eyes
starwars %>%
  filter(homeworld == "Tatooine", eye_color == "brown")
```

```{r filterExercise}
## Filter all women whose mass is unknown
starwars %>%
  filter(sex == 'female',is.na(mass))
```

### `arrange()`

`arrange()` works similarly to `filter()` except that instead of filtering or
selecting rows, it reorders them. It takes a data frame, and a set of column
names (or more complicated expressions) to order by. If you provide more than
one column name, each additional column will be used to break ties in the values
of preceding columns:

```{r arrange}
## Arrange characters by height and mass
starwars %>%
  arrange(height, mass)
```

Use desc() to order a column in descending order:

```{r arrangeDesc}
## Arrange characters by height in descending order
starwars %>%
  arrange(desc(height))
```

```{r arrangeExercise}
## Show height of only women in descending order
starwars %>%
  filter(sex == 'female') %>%
  arrange(desc(height))

## The second option, however, I would say less readable
arrange(filter(starwars, sex == 'female'), desc(height))
```

### `slice()`

`slice()` lets you index rows by their (integer) locations. It allows you to
select, remove, and duplicate rows. We can get characters from row numbers 5
through 10.

```{r slice}
## Select lines from 5 to 10
starwars %>%
  slice(5:10)
```

It is accompanied by a number of helpers for common use cases:

* `slice_head()` -- selects n rows from the top
* `slice_tail()` -- select n rows from the bottom
* `slice_sample()` -- selects n random rows
* `slice_min()` -- selects rows with the lowest value of variable
* `slice_max()` -- selects rows with the highest value of variable

```{r sliceExercie}
## Show in descending order the height of random sample of 10 observations
```

### `select()`

Often you work with large data sets with many columns but only a few are actually
of interest to you. `select()` allows you to rapidly zoom in on a useful subset
using operations that usually only work on numeric variable positions:

```{r select}
## Select hair_color, skin_color and eye_color columns only
starwars %>%
  select(hair_color, skin_color, eye_color)

## Select all columns between hair_color and eye_color
starwars %>%
  select(hair_color:eye_color)

## Select all columns that are not between hair_color and eye_color
starwars %>%
  select(!(hair_color:eye_color))

## Select all columns that contain word color
starwars %>%
  select(ends_with("color"))
```

There are a number of helper functions you can use within `select()`, like
`starts_with()`, `ends_with()`, `matches()` and `contains()`. These let you
quickly match larger blocks of variables that meet some criterion. See `?
select` for more details.

You can rename variables with select() by using named arguments:
```{r selectRename}
## Select homeworld and change its name to homeworld
starwars %>%
  select(home_world = homeworld)
```

But because select() drops all the variables not explicitly mentioned, it’s not
that useful. Instead, use rename():

```{r rename}
## Rename home_world to homeworld
starwars %>%
  rename(home_world = homeworld)
```

```{r renameExercise}
## Rename all American spelling variables to British spelling
## Select all renamed variables and homeland
## Filter only observations with skin color red
## You might want to investigate function grepl() beforehand
```

### `mutate()`

Besides selecting sets of existing columns, it’s often useful to add new columns
that are functions of existing columns. This is the job of mutate():

```{r mutate}
## Create variable height in meters change the order of columns
starwars %>%
  mutate(height_m = height / 100) %>%
  select(height_m, height, everything())
```

```{r mutateExercise}
## For observations that have a color of hair create a new variable
## that is the multiplication of height in meters and mass in kg
```

### `group_by()`

group_by() takes an existing tbl and converts it into a grouped tbl where
operations are performed "by group". It is almost never used separately but
rather before either `mutate()` or `summarise()`.

### `summarise()`

```{r summarise}
## Compute average characters' height in meters for each homeworld
starwars %>%
  group_by(homeworld) %>%
  summarise(height = mean(height)/100)
```

```{r summariseExercise}
## Compute average height for a specie on a given planet
```

```{r summariseExercise2}
## Compute number of characters living on a given planet
## Investigate n() function beforehand
```

### Recode variables

Last time we used a named vector to recode variables. Using the `tidyverse` you
might still use the same approach to recode variables but there are two
fundamental functions that will help you to define a new variable that depends
on the values of the other variable. It might sound confusing at first but let's
imagine that you want to change the value `NA` in `hair_color` to `none`
(although you realize that we will lose some information, right?). There is a
very easy way to do so in `dplyr`. We will use function `if_else()`. It is a
`dplyr` wrapper of `if-statement`. Please check the details by typing into
console `?if_else` to better understand what it does. In general, it takes as
the first argument the logical test, as the second argument is the value when
the if statement is true, and the third argument is the value if the condition
is not fulfilled.

```{r ifElse}
## Recode NA into 'none' in hair_color
starwars %>%
  mutate(hair_color = if_else(is.na(hair_color), 'none', hair_color))
```

The second very useful function is `case_when()` it allows for more than two
options when we want to recode a variable. Imagine that we want to check who is
eligible for roller-coaster. The boundary conditions are that the character must
be tall enough and also old enough. However, we would also like to know which
criteria is met.

```{r caseWhen}
## Create a variable based on both age and height
## If birth_year > 18 and height > 170 than roller-coaster = eligible
## If birth_year > 18 and height < 170 than roller-coaster = adult
## If birth_year > 18 and height > 170 than roller coaster = tall
starwars %>%
  mutate(rollercoster = case_when( birth_year > 18 ~ 'adult',
                                   height > 170 ~ 'tall',
                                   birth_year > 18 && height > 170 ~ 'eligible'))
```

**IMPORTANT:** please note that the order matters. If we first test for
conjunction and afterward for height and age we would never find anyone eligible
for the roller-coaster.

```{r recode}
## Create a new variable BMI (body mass index). Compute BMI for each character
## following the formula from the Wikipedia webpage
## Based on the categorical variable BMI_norms filter only characters that 
## have BMI in the normal range
```

## Data visualization with `ggplot`

Before we start talking about networks and network analysis let's talk about
data visualization in general. It is often the most important part of data
analysis because it allows you to observe what really is happening in your data
set #1pictureIsWorth1000words. In **R**, we have multiple systems for data
visualization. Here, we will focus on the approach that uses similar logic to
the data manipulation we showed you above.

### Data set

For data visualization, we will use a built-in data set that contains
observations collected by the US Environmental Protection Agency on 38 models of
car.

```{r loadData}
mpg
```

Among the variables in mpg are:

1. `displ` -- a car’s engine size, in liters.
2. `hwy` -- a car’s fuel efficiency on the highway, in miles per gallon (mpg). A
car with a low fuel efficiency consumes more fuel than a car with a high fuel
efficiency when they travel the same distance.

Let’s use our first graph to answer a question: **Do cars with big engines use
more fuel than cars with small engines?** You probably already have an answer, but
try to make your answer precise. What does the relationship between engine size
and fuel efficiency look like? Is it positive? Negative? Linear? Nonlinear?

```{r firstPlot}
## Scatter plot displ vs hwy
mpg %>%
ggplot() + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

The graph is probably exactly what we were expecting but how did it happen?
Let's have a closer look. First, we took our data set and passed it to the
function called `ggplot()`. It created a coordinate system to which you could
add layers. Second, we added a layer of points with the use of `geom_point()`
function. In the mapping argument, we specified what should be on x-axis and
what on y-axis

You can add a third variable, like class, to a two-dimensional scatter plot by
mapping it to an anesthetic. Anesthetic is a visual property of the objects in
your plot. Aesthetics include things like the size, shape, or color of your
points. You can display a point (like the one below) in different ways by
changing the values of its aesthetic properties. Since we already use the word
“value” to describe data, let’s use the word “level” to describe aesthetic
properties.

You can convey information about your data by mapping the aesthetics in your
plot to the variables in your data set. For example, you can map the colors of
your points to the class variable to reveal the class of each car.

```{r colorPlot}
## Scatter plot displ vs hwy, with class colored
mpg %>%
  ggplot() + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

```{r shapePlot}
## Scatter plot displ vs hwy, shape indicates the class
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

```{r plotExercise}
## Plot scatter plot dspl vs hwy but the color type of transmission
## Ideally we would have only two types of transmission: manual and auto
## HINT: You might want to create a new variable for that purpose
```


#### Facets

Facets

One way to add additional variables is with aesthetics. Another way,
particularly useful for categorical variables, is to split your plot into
facets, subplots that each display one subset of the data.

To facet your plot by a single variable, use facet_wrap(). The first argument of
`facet_wrap()` should be a formula, which you create with ~ followed by a variable
name (here “formula” is the name of a data structure in R, not a synonym for
“equation”). The variable that you pass to facet_wrap() should be discrete.

```{r facets}
## Scatter plot dipl vs hwy, separeta graphs for manufacturers
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ manufacturer, nrow = 2)
```

To facet your plot on the combination of two variables, add facet_grid() to your
plot call. The first argument of facet_grid() is also a formula. This time the
formula should contain two variable names separated by a ~.

```{r grids}
## Scatter plot
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```

#### Tend Curves

To change the geom in your plot, change the geom function that you add to
`ggplot()`. For instance, instance if you are more interested in the trend curve
than the scatter plot you might use `geom_smooth()`.

```{r trendCurve}
## Trend curve for the relationship between displ and hwy
mpg %>%
  ggplot() + 
    geom_smooth(mapping = aes(x = displ, y = hwy))
```

Every geom function in ggplot2 takes a mapping argument. However, not every
aesthetic works with every geom. You could set the shape of a point, but you
couldn’t set the “shape” of a line. On the other hand, you could set the line
type of a line. geom_smooth() will draw a different line, with a different line
type, for each unique value of the variable that you map to line type.

```{r trendCurveShape}
## Trend curves for the relationship between displ and hwy, linetype indicates drv
mpg %>%
  ggplot() + 
    geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))
```

To display multiple geoms in the same plot, add multiple geom functions to ggplot():

```{r multipleGeoms}
## Combine scatter plot and trend curve
mpg %>%
  ggplot(mapping = aes(x = displ, y = hwy)) + 
    geom_point() + 
    geom_smooth()
```

You can also add a third variable similar to the one we used before.
```{r multiple_geoms_colors}
## Combine scatter plot and trend curve
## Color in scatter plot indicates class
mpg %>%
  ggplot(mapping = aes(x = displ, y = hwy)) + 
    geom_point(mapping = aes(color = class)) + 
    geom_smooth()
```

## Write out data

This is all very good but the question arises how one can somehow save the data
set we spent so much time on? And also how we can import some data to `RStudio`.
Obviously, there are specific functions to do so. Let's start with writing out a
`data.frame` to a file.

```{r irisWriteOut}
## These are global variables and should be defined at the beginning of the notebook!
HOME_PATH <- here::here()
DATA_PATH <- file.path(HOME_PATH, 'data')

## Write out data set to a csv file
iris %>% 
  write.csv2(file = file.path(DATA_PATH,'iris.csv'))
```

```{r plotWriteOut}
## Assign plot to the object plot
plt <- mpg %>%
  ggplot() + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))

## Global variable 
FIG_PATH <- file.path(HOME_PATH, 'fig')

## Write out the plot to the file
plt %>%
  ggsave(filename = file.path(FIG_PATH,"scatter_plot.png"),
       device = "png",
       dpi = 300,
       height = 4.5,
       width = 6)
```

```{r read}
## Read the data from the file
iris_read <- read.csv2(file = file.path(DATA_PATH, 'iris.csv'))
```
